Technique for \earlywarning\ filtering for compact binary coalescence and its implications for multi-messenger astronomy

### 2011-06-28 introduction.tex
lloid_paper: polished introduction; not much change in material.
git-svn-id: https://svn.ligo.caltech.edu/svn/6th-Floor-Papers/lloid_paper@1174 ccdd05d5-56c1-461a-bf98-85f6876523b4

As a compact binary system loses energy to gravitational waves (\GW{}s), its
orbital separation decays, leading to a runaway inspiral with the \GW{}
amplitude and frequency increasing until the system eventually merges.  It is
thought that if a \textsc{ns} is involved, it may become tidally disrupted near
the merger and fuel a bright electromagnetic (\EM{}) counterpart
\citep{shibata:2007}.  The gravitational wave signal is in principle detectable
before merger.  Effort from both the \GW{} and astronomy communities may make
it possible to use \GW{} observations as an early warning trigger for \EM{}
followup.  We have the ambition of reporting GW candidates not minutes after
the merger, as has already been accomplished, but seconds before.  We discuss
one essential ingredient in this effort, a computationally inexpensive
real-time filtering algorithm for detecting inspiral signals in gravitational
wave data.  We also present the prospects for advanced \GW\ detectors and
discuss other areas of work that would be required for rapid analysis.

In the first generation of ground-based laser interferometers, the \GW{}
community initiated a project to send alerts when potential \GW{} transients
were observed in order to trigger followup observations by \EM{} telescopes.
The typical latencies were 30 minutes \citep{HugheyGWPAW2011}, which was an
important achievement, but probably too late to catch the potential prompt
\EM{} emission.

@@

Though compact binary coalescence (\CBC{}) events may be responsible for
several classes of \EM{} transients, \CBC{}s are in particular believed to be a
mechanism for short gamma-ray bursts (short \GRB{}s) \citep{Lee:2005, nakar07}.
In this scenario, prompt gamma-ray emission arises on the accretion timescale
of a central compact object formed after the merger, 0.1--1.0\,s
\citep{Janka1999}.  Interestingly, roughly 1/4 -- 1/2 of observed short \GRB{}s
also exhibit extended emission $\sim100$\,s.  Potential explanations for the
emission are delayed fall-back accretion \citep{Metzger2010} caused by
$r$-process heating or the formation of a proto-magnetar that interacts with
ejecta \citep{Bucciantini2011}.  In the case of the proto-magnetar, it might be
formed from accretion induced collapse, or the merger of two neutron stars.  An
accompanying \GW{} observation would confirm the latter.

Optical flashes have only been observed for a handful of long \GRB{}s
\citep{2011CRPhy..12..255A} by telescopes with extremely rapid response or, in
the case of \textsc{grb 080319b}, by pure serendipity, where several telescopes
were already observing the afterglow of another \GRB{} in the same field of
view~\citep{2008Natur.455..183R}. The observed optical flashes peaked within
tens of seconds and decayed quickly.  Short \GRB{}s, on the other hand,
typically fade too quickly to even catch the tails of the afterglows in any
band. Rapid \GW{} transient alerts could enable the first observation of prompt
optical flashes and the rise of afterglows from short \GRB{}s, and, for a truly
exceptional event, perhaps even a glimpse of a precursor
\citep{0004-637X-723-2-1711}.

@@

\begin{figure}[h]
\plotone{figures/snr_in_time}
\caption{\label{fig:earlywarning}Expected number of \textsc{ns}--\textsc{ns}
sources that could be detectable by Advanced \LIGO\ a given number of seconds
before coalescence.  The heavy solid line is the most realistic yearly rate
estimate.  The shaded region represents the 5 to 95\% confidence interval
arising from uncertainty in predicted event rates \citep{Abadie:2010p10836}.}
\end{figure}
%
Examining the signal to noise ratio (\SNR{}) stream for threshold crossings
before the \GW{} signal leaves the detection band makes it possible to trade
some \SNR{} and sky localization accuracy for pre-merger detection.
Figure~\ref{fig:earlywarning} shows projected early detectability rates for
\textsc{ns}--\textsc{ns} binaries in Advanced \LIGO\ assuming event rates
predicted in \citet{Abadie:2010p10836} and anticipated detector sensitivity for
the `zero detuning, high power' configuration described in \citet{ALIGONoise}.
The most realistic estimates indicate that at an \SNR\ threshold of 8, we will
observe a total of 40~events~yr$^{-1}$. $\sim$10~yr$^{-1}$ will be detectable
within 10~s of merger and $\sim$5~yr$^{-1}$ will be detectable within 25~s of
merger if analysis can proceed with near zero latency. These rates are shown
with their substantial uncertainties in Figure~\ref{fig:earlywarning}.

@@

\begin{table}[h]
\caption{\label{table:sky-localization-accuracy}Horizon distance, \SNR\ at
merger, and area of 90\% confidence at selected times before merger for sources
with expected detectability rates of 40, 10, 1, and 0.1~yr$^{-1}$.}
\begin{center}
\begin{tabular}{rrrrrrr}
\tableline\tableline
rate & horiz. & final & \multicolumn{4}{c}{$A$(90\%) (deg$^2$)} \\
\cline{4-7}
yr$^{-1}$ & (Mpc) & \SNR\ & 25 s & 10 s & 1 s & 0 s \\
\tableline
40\phd\phn & 445 & 8.0 & 15500 & 3100 & 200 & 9.6 \\
10\phd\phn & 280 & 12.7 & 6100 & 1200 & 78 & 3.8 \\
1\phd\phn & 130 & 27.4 & 1300 & 260 & 17 & 0.8 \\
0.1 & 60 & 58.9 & 280 & 56 & 3.6 & 0.2 \\
\tableline
\end{tabular}
\end{center}
\end{table}
%
\begin{figure}[h]
\plotone{figures/loc_in_time}
\caption{\label{fig:sky-localization-accuracy}Area of the 90\% confidence
region as a function of time before coalescence for sources with anticipated
detectability rates of 40, 10, 1, and 0.1~yr$^{-1}$. The heavy dot indicates
the time at which the accumulated \SNR\ exceeds a threshold of 8.}
\end{figure}

Before discussing realistic latencies, it is important to note that \EM{}
followup requires estimating the location of the \GW{} source. The localization
uncertainty is dominated by uncertainty in the time of arrival of the \GW{}s,
which is determined by the signal's effective bandwidth and \SNR{}
\citep{Fairhurst2009}.  Table~\ref{fig:earlywarning} and
Figure~\ref{fig:sky-localization-accuracy} show the estimated minimum 90\%
confidence area versus time of the loudest coalescence events detectable by
Advanced \LIGO{} and Advanced Virgo.  Once per year, we expect to observe an
event with a final \SNR\ of $\sim 27$ whose location can be constrained to about
1300~deg$^2$ (3.1\% of the sky) using only the waveform 25~s before merger,
260~deg$^2$ (0.63\% of the sky) using the waveform 10~s before merger, and
0.82~deg$^2$ (0.0020\% of the sky) using the full \GW\ waveform.

@@

Here we provide a brief list of technical latencies associated with \CBC\
analysis in S6, VSR3 in chronological order \citep{HugheyGWPAW2011}.
%
\begin{itemize}
%
\item {\it raw data aggregation and broadcasting ($sim 0.25$\,s):}  There are
fundamental latencies associated with data acquisition and broadcasting that
cannot be overcome.  For example, the \LIGO{} data acquisition records science
data in 1/16~s blocks~\citep{Bork2001}. Data must also be aggregated from all
of the \GW\ observatories using global computer networks capable of high
bandwidth but perhaps only modest latency.  This could introduce an additional
latency of $\sim$100~ms.  These latencies are microscopic compared with the
other technical sources of latency below. 
%
\item {\it Data calibration, quality control and distribution ($\sim
2$\,minutes):}  This is the time it took to calibrate the detector data, assess
its quality and distribute it to a central location for analysis.  Improvements
in other areas could make this the dominating source of latency. If possible
work should be done to improve this.
%
\item {\it Trigger production ($2-5$\,minutes):} This was an impressive latency
and was certainly sub-dominant in the S6, VSR3 analysis.  However, even if no
other sources of latency exist it is too long to catch prompt or extended
emission.  It is also more challenging for searches with advanced detectors
because the templates needed to filter the data are about 10 times longer.  We
will focus on reducing the latency of the trigger production step in this work.
%
\item {\it Event gathering, downselection and pointing calculations ($2-3$\,
minutes):} This was the time required to gather the event parameters and sky
localization from the various online analyses, downselect the events and
calculate pointings.  If other sources of latency were improved this could
dominate.  If possible work should be done to improve this.
%
\item {\it Human vetting of the trigger quality ($10-20$\,minutes):}  This was
by far the largest latency during the S6, VSR3 low latency analysis and will
hopefully be automated by the advanced detector era.  We therefore give it no
more consideration here.
%
\end{itemize}
%

This work will focus on reducing the latency of trigger production.  Realizing
advance detection of \CBC{}s will require striking a balance between latency
and throughput of filtering \GW\ wave data. \CBC{} searches consist of banks of
matched filters, or cross-correlations between the data stream and a bank of
nominal ``template'' signals.  There are many different implementations of
matched filters, but most have high throughput at the cost of high latency, or
low latency at the cost of low throughput.  The former are epitomized by the
overlap-save algorithm~% \citep{numerical-recipes-chapter-13}
for frequency domain (\FD) convolution, currently the preferred method in \GW{}
searches.  The most obvious example of the latter is the time domain (\TD)
convolution, which is latency-free.  However, its computational complexity is
quadratic in the length of the templates, so it is prohibitively expensive for
long templates.

>>>>

As a compact binary system loses energy to gravitational waves (\GW{}s), its
orbital separation decays, leading to a runaway inspiral with the \GW{}
amplitude and frequency increasing until the system eventually merges.  It is
thought that if a neutron star (\textsc{ns}) is involved, it may become tidally disrupted near
the merger and fuel a bright electromagnetic (\EM{}) counterpart
\citep{shibata:2007}.  The gravitational-wave signal is in principle detectable
before merger.  Effort from both the \GW{} and astronomy communities may make
it possible to use \GW{} observations as an early warning trigger for \EM{}
followup. In the first generation of ground-based laser interferometers, the \GW{}
community initiated a project to send alerts when potential \GW{} transients
were observed in order to trigger followup observations by \EM{} telescopes.
The typical latencies were 30 minutes \citep{HugheyGWPAW2011}, which was an
important achievement, but too late to catch the potential prompt \EM{}
emission. We have the ambition of reporting GW candidates not minutes after
the merger, as has already been accomplished, but seconds before.  We discuss
one essential ingredient in this effort, a computationally inexpensive
real-time filtering algorithm for detecting inspiral signals in gravitational
wave data.  We also present the prospects for advanced \GW\ detectors and
discuss other areas of work that would be required for rapid analysis.

@@

Though compact binary coalescence (\CBC{}) events may be responsible for
several classes of \EM{} transients, \CBC{}s are in particular believed to be a
mechanism for short gamma-ray bursts (short \GRB{}s) \citep{Lee:2005, nakar07}.
In this scenario, prompt gamma-ray emission arises on the accretion timescale
of a central compact object formed after the merger, 0.1--1.0\,s
\citep{Janka1999}.  Interestingly, roughly a quarter to half of observed short
\GRB{}s
also exhibit extended emission $\sim$$100$\,s.  Potential explanations for the
emission are delayed fall-back accretion \citep{Metzger2010} caused by
$r$-process heating or the formation of a proto-magnetar that interacts with
ejecta \citep{Bucciantini2011}.  In the case of the proto-magnetar, it might be
formed from accretion induced collapse, or the merger of two \textsc{ns}s.  An
accompanying \GW{} observation would confirm the latter.

Optical flashes have only been observed for a handful of long \GRB{}s
\citep{2011CRPhy..12..255A} by telescopes with extremely rapid response or, in
the case of \textsc{grb 080319b}, by pure serendipity, where several telescopes
were already observing the afterglow of another \GRB{} in the same field of
view \citep{2008Natur.455..183R}. The observed optical flashes peaked within
tens of seconds and decayed quickly.  Short \GRB{}s, on the other hand,
typically fade too quickly to even catch the tails of the afterglows in any
band. Rapid \GW{} transient alerts could enable the first observation of prompt
optical flashes and the rise of afterglows from short \GRB{}s.

@@

\begin{figure}[h]
\plotone{figures/snr_in_time}
\caption{\label{fig:earlywarning}Expected number of \textsc{ns}--\textsc{ns}
sources that could be detectable by Advanced \LIGO\ a given number of seconds
before coalescence.  The heavy solid line is the most realistic yearly rate
estimate.  The shaded region represents the 5 to 95\% confidence interval
arising from uncertainty in predicted event rates \citep{Abadie:2010p10836}.}
\end{figure}
%
Examining the signal to noise ratio (\SNR{}) stream for threshold crossings
before the \GW{} signal leaves the detection band makes it possible to trade
some \SNR{} and sky localization accuracy for pre-merger detection.
Figure~\ref{fig:earlywarning} shows projected early detectability rates for
\textsc{ns}--\textsc{ns} binaries in Advanced \LIGO\ assuming event rates
predicted in \citet{Abadie:2010p10836} and anticipated detector sensitivity for
the `zero detuning, high power' configuration described in \citet{ALIGONoise}.
The most realistic estimates indicate that at an \SNR\ threshold of 8, we will
observe a total of 40\,events\,yr$^{-1}$. $\sim$10\,yr$^{-1}$ will be detectable
within 10\,s of merger and $\sim$5\,yr$^{-1}$ will be detectable within 25\,s of
merger if analysis can proceed with near zero latency. These rates are shown
with their substantial uncertainties in Figure~\ref{fig:earlywarning}.

@@

\begin{table}[h]
\caption{\label{table:sky-localization-accuracy}Horizon distance, \SNR\ at
merger, and area of 90\% confidence at selected times before merger for sources
with expected detectability rates of 40, 10, 1, and 0.1\,yr$^{-1}$.}
\begin{center}
\begin{tabular}{rrrrrrr}
\tableline\tableline
rate & horiz. & final & \multicolumn{4}{c}{$A$(90\%) (deg$^2$)} \\
\cline{4-7}
yr$^{-1}$ & (Mpc) & \SNR\ & 25 s & 10 s & 1 s & 0 s \\
\tableline
40\phd\phn & 445 & 8.0 & 15500 & 3100 & 200 & 9.6 \\
10\phd\phn & 280 & 12.7 & 6100 & 1200 & 78 & 3.8 \\
1\phd\phn & 130 & 27.4 & 1300 & 260 & 17 & 0.8 \\
0.1 & 60 & 58.9 & 280 & 56 & 3.6 & 0.2 \\
\tableline
\end{tabular}
\end{center}
\end{table}
%
\begin{figure}[h]
\plotone{figures/loc_in_time}
\caption{\label{fig:sky-localization-accuracy}Area of the 90\% confidence
region as a function of time before coalescence for sources with anticipated
detectability rates of 40, 10, 1, and 0.1\,yr$^{-1}$. The heavy dot indicates
the time at which the accumulated \SNR\ exceeds a threshold of 8.}
\end{figure}

Before discussing realistic latencies, it is important to note that \EM{}
followup requires estimating the location of the \GW{} source. The localization
uncertainty is dominated by uncertainty in the time of arrival of the \GW{}s,
which is determined by the signal's effective bandwidth and \SNR{}
\citep{Fairhurst2009}.  Table~\ref{fig:earlywarning} and
Figure~\ref{fig:sky-localization-accuracy} show the estimated minimum 90\%
confidence area versus time of the loudest coalescence events detectable by
Advanced \LIGO{} and Advanced Virgo.  Once per year, we expect to observe an
event with a final \SNR\ of $\sim 27$ whose location can be constrained to about
1300\,deg$^2$ (3.1\% of the sky) using only the waveform 25\,s before merger,
260\,deg$^2$ (0.63\% of the sky) using the waveform 10\,s before merger, and
0.82\,deg$^2$ (0.0020\% of the sky) using the full \GW\ waveform.

@@

There were a number of sources of latency associated with \CBC{}
analysis in S6/VSR3 \citep{HugheyGWPAW2011}, listed here in chronological order.
%
\begin{itemize}
%
\item {\it Data aggregation and broadcasting ($\sim$$0.25$\,s):}
The \LIGO{} data acquisition system records science
data in 1/16\,s blocks \citep{Bork2001}. Data must also be aggregated from all
of the \GW\ observatories using global computer networks capable of high
bandwidth but perhaps only modest latency.  This could introduce an additional
latency of $\sim$100\,ms.  These latencies are much smaller than the
other sources of latency below and would require significant reengineering and
capital investments to overcome.
%
\item {\it Data calibration, quality control, and distribution
($\sim$$2$\,min):} There is technical latency in calibrating the detector data,
assessing its quality, and distributing it to a central location for analysis.
None of this latency is fundamental and improvements in other areas could make
this the dominating source of latency, so work should be done to improve this.
%
\item {\it Trigger production ($2$--$5$\,min):} This was an impressive latency
and was certainly sub-dominant in the S6/VSR3 analysis.  However, even if no
other sources of latency exist, triggering latency is too long to catch prompt
or extended
emission. It is also more challenging for searches with advanced detectors
because the templates needed to filter the data are about ten times longer.  We
will focus on reducing the latency of trigger production in this work.
%
\item {\it Event gathering, downselection, and pointing calculations
($2$--$3$\,min):} S6/VSR3 saw the introduction of low-latency astronomical
alerts, which required gathering event parameters and sky localization from the
various online analyses, downselecting the events, and calculating pointings.
If other sources of latency improve, this could dominate, so work should be
done to improve this.
%
\item {\it Human vetting of the trigger quality ($10$--$20$\,min):} In
commissioning the new alert system in S6/VSR3, humans manually examined each
candidate for obvious flaws before publishing an alert. This was by far the
largest latency during S6/VSR3 and hopefully confidence will grow to the
point where human operators are satisfied with checks after alerts have been
transmitted. We give it no further consideration here.
%
\end{itemize}

This work will focus on reducing the latency of trigger production.  Realizing
advance detection of \CBC{}s will require striking a balance between latency
and throughput of filtering \GW\ data. \CBC{} searches consist of banks of
matched filters, or cross-correlations between the data stream and a bank of
nominal ``template'' signals.  There are many different implementations of
matched filters, but most have high throughput at the cost of high latency, or
low latency at the cost of low throughput.  The former are epitomized by the
overlap-save algorithm % \citep{numerical-recipes-chapter-13}
for frequency domain (\FD) convolution, currently the preferred method in \GW{}
searches.  The most obvious example of the latter is the time domain (\TD)
convolution, which is latency-free.  However, its computational complexity is
quadratic in the length of the templates, so it is prohibitively expensive for
long templates.
